---
title: "DMA"
output: html_document
---

# Differential Methylation Analysis (GSE111629; Parkinson's disease)

```{r}
rm(list=ls())
gc(full=T)
```

```{r}
BiocManager::install("pheatmap")
```

```{r}
library(limma)
library(ggplot2)
library(pheatmap)
library(tidyverse)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
```

### 1. Setup design matrix

**Design matrix includes apart from biological gender and age also cell proportions on methylation probes as covariants.**

```{r}
targets <- readRDS("targets_final.rds")
targets$`age:ch1` <- as.numeric(targets$`age:ch1`)
targets$`gender:ch1` <- as.character(targets$`gender:ch1`)
names(targets)[names(targets) == 'gender:ch1'] <- 'Sex'
names(targets)[names(targets) == 'age:ch1'] <- 'Age'
design <- model.matrix(~0 + Sample_Group + 
                       Sex + 
                       Age + 
                       CD8T + CD4T + NK + Bcell + Mono, 
                       data = targets)

print("Design matrix columns:")
print(colnames(design))
```

### 3. Check whether covariants correlate up to a collinear degree

```{r}
cor_matrix <- cor(design[, -1])
heatmap(cor_matrix, 
        main = "Covariate Correlation Matrix",
        margins = c(10, 10))
```

### 4. Perform differential analysis on methylation probes

```{r}
m_values <- readRDS("m_values_final.rds")
fit <- lmFit(m_values, design)
cont.matrix <- makeContrasts(Parkinsons_vs_Control = Sample_GroupPD - Sample_GroupControl, levels = design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
results <- topTable(fit2, number = Inf, coef = "Parkinsons_vs_Control")
print("Sum of statistically significant differentially methylated probes")
sum(results$adj.P.Val < 0.05 & abs(results$logFC) > 0.2)
```

### 5. Annotate the differentially methylated probes by using the respective 450k methylation array annotation package

```{r}
results$ProbeID <- rownames(results)
ann_450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
ann_sub <- ann_450k[rownames(results), c("chr", "pos", "UCSC_RefGene_Name", "UCSC_RefGene_Group", "Relation_to_Island")]
annotated_results <- merge(results, ann_sub, by.x = "ProbeID", by.y = "row.names")
annotated_results <- annotated_results[order(annotated_results$adj.P.Val, -annotated_results$logFC), ]
write.csv(annotated_results, "differential_methylation_results.csv", row.names = FALSE)
```

```{r}
if (!exists("annotated_results")) {
  annotated_results <- read.csv("differential_methylation_results.csv", )
}
```

### 6. Visualize the results in a volcano plot

*Thresholds for fold change on M-values and statistical significance is set to absolute 0.2 and 0.05 respectively. The resulting subset is still subject for Δβ value assessment to estimate biological significance.*

```{r}
volcano_plot <- ggplot(annotated_results, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(alpha = 0.6, aes(color = (adj.P.Val < 0.05 & abs(logFC) > 0.2))) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-0.2, 0.2), linetype = "dashed") +
  ggtitle("Volcano Plot of Differential Methylation") +
  theme_minimal()
print(volcano_plot)
```

### 7. Visualize beta values of the differentially methylated probes in a heatmap using the top 10 CpG's.

*Here a scale is used instead of absolute values*

```{r}
beta_matrix <- readRDS("beta_matrix_final.rds");
# write.csv(beta_matrix, "./beta_matrix_final.csv")
annotated_filtered <- as.data.frame(annotated_results) %>% filter(abs(logFC) > 0.2 & adj.P.Val < 0.05)
top_probes <- annotated_filtered$ProbeID[1:10]
top_beta <- beta_matrix[rownames(beta_matrix) %in% top_probes, ]

annotation_col <- data.frame(Group = targets$Sample_Group)
rownames(annotation_col) <- colnames(top_beta)

pheatmap(top_beta,
         annotation_col = annotation_col,
         show_rownames = T,
         show_colnames = F,
         main = "Top 10 Differentially Methylated Probes",
         scale = "row")
```

### 8. Visualize *absolute* beta values of the differentially methylated probes in a heatmap using the top 10 CpG's.

```{r}
pheatmap(top_beta,
         annotation_col = annotation_col,
         show_rownames = T,
         show_colnames = F,
         main = "Top DMPs - Absolute Beta Values",
         scale = "none",
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

### 9. Transform beta matrix row names to match with metadata from targets

```{r}
beta_matrix_filtered <- beta_matrix[rownames(beta_matrix) %in% annotated_filtered$ProbeID, ]
beta_matrix_t <- as.data.frame(t(beta_matrix_filtered))
beta_matrix_t <- beta_matrix_t %>%
  rownames_to_column("row_id") %>%
  mutate(before_underscore = sub("_.*", "", row_id)) %>%
  column_to_rownames("before_underscore")
beta_matrix_t$Sample_Group <- targets$Sample_Group[match(rownames(beta_matrix_t), targets$Sample_Name)]
beta_matrix_t$Gender <- targets$Sex[match(rownames(beta_matrix_t), targets$Sample_Name)]
beta_matrix_t$CD8T <- targets$CD8T[match(rownames(beta_matrix_t), targets$Sample_Name)]
beta_matrix_t$CD4T <- targets$CD4T[match(rownames(beta_matrix_t), targets$Sample_Name)]
beta_matrix_t$NK <- targets$NK[match(rownames(beta_matrix_t), targets$Sample_Name)]
beta_matrix_t$Bcell <- targets$Bcell[match(rownames(beta_matrix_t), targets$Sample_Name)]
beta_matrix_t$Mono <- targets$Mono[match(rownames(beta_matrix_t), targets$Sample_Name)]
write.csv(beta_matrix_t, "./beta_matrix_filtered_meta.csv")
```

### 10. Calculate means per CpG probe for each sample group and calculate Δβ values

```{r}
pd_samples <- rownames(beta_matrix_t[beta_matrix_t$Sample_Group == 'PD', ])
control_samples <- rownames(beta_matrix_t[beta_matrix_t$Sample_Group == 'Control', ])

beta_matrix_meta <- t(beta_matrix_t[!names(beta_matrix_t) %in% c("row_id", "Sample_Group", "Gender", "CD8T", "CD4T", "NK", "Bcell", "Mono")])

colnames_pd <- which(colnames(beta_matrix_meta) %in% pd_samples)
colnames_control <- which(colnames(beta_matrix_meta) %in% control_samples)

mean_beta_pd <- rowMeans(beta_matrix_meta[, colnames_pd], na.rm = TRUE)
mean_beta_control <- rowMeans(beta_matrix_meta[, colnames_control], na.rm = TRUE)

delta_beta <- mean_beta_pd - mean_beta_control

annotated_filtered$mean_beta_control <- mean_beta_control
annotated_filtered$mean_beta_pd <- mean_beta_pd
annotated_filtered$delta_beta <- delta_beta
rownames(annotated_filtered) <- annotated_filtered$ProbeID
annotated_filtered <- annotated_filtered[-1]
write.csv(annotated_filtered, "annotated_delta_beta.csv", )
```

### 11. Filter probes that are below a particular Δβ-threshold

**The threshold is based on researcg evidence in particular with analysing whole-blood methylation data for PD. Since Δβ-values of a great magnitude aren't expected in samples expected to reflect a neurodegenerative disease (opposed to cancers samples where such values might emerge as high as 20%) a very low threshold is set to 3% while other studies also go as low as 2% (Latapi et al. 2024).**

> Gonzalez-Latapi, P., Bustos, B., Dong, S., Lubbe, S., Simuni, T., & Krainc, D. (2024). Alterations in Blood Methylome as Potential Epigenetic Biomarker in Sporadic Parkinson’s Disease. Annals of Neurology, 95(6), 1162–1172. <https://doi.org/10.1002/ANA.26923>

```{r}
annotated_filtered_delta_beta <- annotated_filtered[abs(annotated_filtered$delta_beta) > 0.03, ]
write.csv(annotated_filtered_delta_beta, file="annotated_filtered_delta_beta.csv")
```
